What does it take to create with domain-appropriate tools?

### 2020-01-15 convivial-2020/tex/manuscript.tex
Abstract criteria

@@

\title{What does it take to create with domain-appropriate tools? A case study on the ``OROM'' system.}
\author{Joel Jakubovic}
\affiliation{University of Kent, Canterbury}

@@

However, this didn't happen. Or at least, it is happening glacially
slowly. The brute fact is that whenever you want to create software, you
go to a text editor and figure out how to translate your design into
that. The text editors, being software, were written with the help of
previous text editors, and so on. It's undeniable that text editors have
improved, even if you think it peaked with Emacs. We just don't seem
able to go beyond them.

\begin{quote}
You discuss what's wrong with text editor in the next few paragraphs,
but I think it would be useful to give some brief summary here so that
readers don't think that ``we need to go beyond text editors'' is
unmotivated - I think the summary of the text below is that you want to
create software using tools that are closer to the domain of the
software (text is far from everything; visual editors would be closer to
a more visual domain) This general formulation might also nicely frame
some of the things you say next.
\end{quote}

Amdahl's Law generalises the following idea: even when you spend hours
of effort doubling the performance of a component used 1\% of the time,
your reward is a system overall improved by a mere 0.5\%.

Text coding is certainly ubiquitous, the 99\% case in programming, so
you might wonder where I'm going with this. Well, a small improvement to
text editing, if adopted by everyone, certainly does have a massive
\emph{intermediate} effect -- but this only \emph{matters} to the extent
that text was helping the programmers in the first place. If my goal is
to draw or animate pictures, or create a digital synth from a frequency
spectrogram, then giving me the ability to auto-indent my SVG markup is
rather underwhelming as a productivity increase, as it doesn't target
the core of the enterprise that makes it so hard.

My experience of coding, most of the projects requiring shapes (such as
GUIs), leads me to conclude that no matter how much I improve my skill
at a particular language, knowledge of libraries or even general coding
ability, my predicament stays the same. Our basic method of creating
software is optimised for an ever-diminishing proportion of the software
we actually want to make; ill-optimised for the graphics, layout,
interactivity and and basic physics -- more on this later -- that we
usually require.

@@

Surely the trick is to \emph{use} coding to build something \emph{better
than it}. And then use that, to build something even better. But there
is an enormous breadth and depth of philosophies here, along with many
failed historical attempts to do better -- or at least, ones that failed
to catch on. But even worse than this is that in my very \emph{language}
here I am making the same mistake as the text editor -- speaking in
unqualified terms of ``better'' and ``worse'' as if there really is a
one-size-fits-all solution to software creation!

Of course what we \emph{really} want is the ability for people to create
\emph{in the way that they think is best} in their particular context --
to equip them to feasibly create the tools that suit them for the thing
they want to make. And second-order tools that suit them for making the
first-order tools, etc. It would do no good to replace text-imperialism
with anything-else-imperialism, which is one common interpretation of
calls for alternatives. If someone wants to type out pictures in ASCII,
let them -- whether they do it for a challenge, or even if they find
that more natural for themselves. But equally, if I want to do it
another way, then please give me that affordance.

\begin{quote}
Is programming like `craft'? I guess some artisans make their own tools,
but most likely not in the higher-order sense you are discussing here.
This suggests that programming is different in some sense - but in what
sense?
\end{quote}

This is how I segue into the software artefact for which I have been
attempting to build a natural representation. True, it is a programmer's
artefact, but it is still representative of what any normal person has
to do, insofar as:

@@

\begin{quote}
I think it would be good to say (fairly generally) what kind of system
do you imagine we should be using (and goals that your OROM/SVG is
getting closer to). This also clarifies why nothing like that exists
already. I suppose you want something that lets you program in a way
that is ``close to the problem domain'' (i.e.~visually if the domain is
visual) and something that has no arbitrary boundaries
(i.e.~meta-circular?). Boxer is the former, but not the latter;
Smalltalk/LISP is the latter but not the former.
\end{quote}

@@

When I first read the paper Open, Reusable Object Models, I was hooked
on its idea of a small but expressive starting system that could be
self-improved into anything. It describes a late-bound, Smalltalk-style
objects and messaging environment that the authors call ``Id'' -- but
which I refer to as OROM, for pronunciation and Googleability.

\begin{quote}
How much does the reader need to know about OROM for the next
paragraphs? I think they need to know there are objects with references
that send messages to each other - this will be clear to those who know
what `Smalltalk-style' means, but not others. It would be useful to give
a very brief summary before getting to the next part of the story.
\end{quote}

The paper as a publication follows the usual patterns. It consists of
mostly prose, several code listings, some diagrams, and full C sources
for a sample implementation at the end. Insofar as this is normal, I
cannot criticise it. But reading this paper, and trying to grasp their
ideas, brought to my attention how inadequate this norm can sometimes
be.

\begin{quote}
The above feels a bit more chatty than necessary - you could probably
shorten the paragraph into just one sentence without losing much.
\end{quote}

In order to understand it in the first place, I repeatedly found myself
drawing extra diagrams. For example, the first acts of the running
system boil down to initialising the three or so objects. This consists
of allocating memory, interpreting it as a C struct and then filling in
fields in a mundane manner. I had great difficulty grasping the
specifics in my head, but when I drew tables in the style of the paper's
diagrams I readily saw what what going on.

Still, areas like messaging semantics could be quite confusing. After
all, should we expect to be able to predict the entire future evolution
of a dynamical system from a static description of its initial state,
such as source code? Is this not why debuggers exist? Having
``de-compiled'' English text and C source code into object diagrams, it
is a shame to have to compile it all back to struct member assignments.

@@

But further than that -- why can't the system be \emph{built} out of
tables in the first place? Shouldn't this be the main takeaway from the
amount of time we spend prototyping, explaining and designing software
as diagrams on paper? Why must the ``natural representation'' be
restricted to the finished product?

@@

\hypertarget{orom-as-html-tables}{%
\subsubsection{OROM as HTML tables}\label{orom-as-html-tables}}

@@

(Describe basic encoding of obj-dicts as tables, use of css class, etc)

Unfortunately, the choice of ordinary HTML as a substrate proved rather
two-edged. I was grateful for the browser's management of graphical
layout, resizable text fields, and keeping the DOM tree synchronised
with what you actually see. This last property enabled me to make the
decision to \emph{directly} encode much of the system state in the DOM,
achieving basic liveness (the thing on the screen is the actual thing)
for the keys and values of obj-dicts.

On the other hand, the browser requires many features to do its job of
rendering complex web pages. And sadly, as its client, I could only make
use of those capabilities which the W3C had decided, at the time of
authorship, were worth the effort exposing in JavaScript. For anything
else, the browser is a black box, and this was very frustrating in the
following case.

\hypertarget{the-radical-concept-of-arrows-that-stay-on-the-shapes}{%
\paragraph{The Radical Concept of Arrows That Stay On The
Shapes}\label{the-radical-concept-of-arrows-that-stay-on-the-shapes}}

@@

Reluctantly, I stuck with my plan B: each object has a numerical ID and
pointers are just fields containing a number, followed using a deref()
function.

@@

\hypertarget{orom-as-files-and-directories}{%
\subsubsection{OROM as files and
directories}\label{orom-as-files-and-directories}}

@@

\hypertarget{orom-as-svg-trees}{%
\subsubsection{OROM as SVG trees}\label{orom-as-svg-trees}}

@@

\hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
\subsection{Inevitable Requirements Of Most Software (and the work we
must do to meet
them)}\label{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}}

@@

\hypertarget{as-a-mere-consumer}{%
\subsubsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}

@@

\hypertarget{retained-mode-vector-graphics}{%
\paragraph{Retained-Mode Vector
Graphics}\label{retained-mode-vector-graphics}}

@@

On this requirement, SVG fits the bill very well. Although it is not
part of the JavaScript language \emph{per se}, it is a standard and
widely supported technology of the Web \emph{platform}. We can observe
that anyone with a browser \emph{in principle} has access to a powerful
vector graphics editor -- just one with no GUI. I will return to this
later.

@@

\hypertarget{basic-assumptions-about-physical-objects}{%
\paragraph{Basic Assumptions About Physical
Objects}\label{basic-assumptions-about-physical-objects}}

@@

All humans learn a basic set of expectations about the things they see
around them. Some of these, such as ``things fall down'', are not
generally appropriate to software UIs -- perhaps because the screen has
a role in our lives a more of a table work surface rather than a
vertical wall, even if it is vertical in real life.

@@

\hypertarget{maintaining-relationships-over-time}{%
\paragraph{Maintaining Relationships Over
Time}\label{maintaining-relationships-over-time}}

@@

\hypertarget{nut-cracking-with-sledgehammers}{%
\paragraph{Nut-Cracking With
Sledgehammers}\label{nut-cracking-with-sledgehammers}}

@@

The inevitable requirements I suggest here, do \emph{not} necessitate
Fully General anything. In fact, such representations would make it more
cumbersome to express what I wanted in OROM. As the old wisdom goes,
there's no point expressing a simple regex search as an arbitrary Turing
machine; my boxes don't \emph{have} a moment of inertia or a mass and I
don't \emph{need} a linear optimisation solver for my space management
-- for the time being.

@@

\hypertarget{as-a-developer}{%
\subsubsection{As A ``Developer''}\label{as-a-developer}}

@@

\hypertarget{practicality-vs.-high-aesthetics}{%
\paragraph{Practicality vs.~High
Aesthetics}\label{practicality-vs.-high-aesthetics}}

@@

(Listing involving CSS, element querying / munging)

and in OROM/SVG the story is similar. It has the advantage of the
powerful developer tools included in browsers, and avoids the need to
manually synchronise separate Model and View for such encoded state.
With a little work, it also allowed the system a large degree of
externalisability -- see Section ??.

@@

Another example of practicality was in the issue of positioning and
sizing. In Section ?? I mentioned that I was able to sidestep some
layout work by offloading it to the user. While building the system I
tended to want to draw boxes into existence, and for that my brain's
existing aesthetic algorithms can decide where they go. There are areas
where the machine must be able to figure this out itself -- I don't want
to be constantly bothered by prompts to place boxes as they are
allocated by some running code -- but for one-offs it is a very
convenient way to do a little less yak-shearing.

@@

\hypertarget{positioning-and-sizing}{%
\paragraph{Positioning and Sizing}\label{positioning-and-sizing}}

@@

\hypertarget{visible-coordinate-systems}{%
\paragraph{Visible Coordinate
Systems}\label{visible-coordinate-systems}}

@@

All child elements of a node transform with it, so already SVG has baked
in a basic facility for translational rigidity. This is only available
as a tree hierarchy, but it is still useful. However, it conflicts with
my early decision to have Point objects all share the global co-ordinate
system. This way, connecting ``this point'' to the mouse pointer, etc.
is not infuriating to express. Still, it was necessary in the case of
certain elements -- especially those which must transcend the tree
structure altogether, like arrows between boxes -- to bite this bullet,
one way or another.

@@

Again, I return to how we tend to work things out in the freedom of
paper. Co-ordinate systems, here merely positionally displaced, have
their origins here and there and have vectors between them. The rods
thus far let me visually express relations between global Points; now
was a question of expressing one global Point as a displacement from
another (the \texttt{\textless{}g\textgreater{}} transform). New rod
Observables p2\_from\_p1 and p1\_from\_p2 do the vector subtraction,
which can then be propagated as local co-ordinates to children. It is
nicer to express the relation (as well as see it!) this way -- see
Figure ?? above.

\hypertarget{context-appropriate-ontologies}{%
\paragraph{Context-appropriate
ontologies}\label{context-appropriate-ontologies}}

\ldots{}

Talk about: fixed ontologies (a rect IS x,y,w,h, and nothing else!)
imposed on programmer

Example from 3.1.1. mention: CSS animations?

``One size fits all'' rears its ugly head again.

I want: top-left, top-right, bot-left, bot-right \ldots{}

Instead of explaining ``the x y is the top left blah blah'' in English
in the W3C spec, they should make these relationships \emph{machine
readable} with information like:

A Rect is

\begin{itemize}
\tightlist
\item
  a Polygon
\item
  defined by 4 degrees of freedom x y width height
\item
  where the four vertices are

  \begin{itemize}
  \tightlist
  \item
    \texttt{{[}x,y{]}} ``top-left''
  \item
    \texttt{{[}x+width,y{]}} ``top-right''
  \item
    etc
  \end{itemize}
\end{itemize}

Reference to paper related to Antranig involving knitting\ldots{}?

\ldots{}

\hypertarget{extensional-functions}{%
\paragraph{Extensional Functions}\label{extensional-functions}}

aka dicts, Maps, std::vector, \ldots{}

\ldots{} \ldots{} Some parts of the OROM authors' C code were confusing
until I realised they were just the guts of a basic associative-array
implementation. \ldots{} \ldots{}

Inevitability / ubiquity of state descriptions that are at trees of
strings with pointers

Antranig ``natural co-ordinates''?

OROM/SVG as Boxer-like extensional functions

\hypertarget{externalisability}{%
\paragraph{Externalisability}\label{externalisability}}

>>>>

\lstdefinelanguage{JavaScript}{
  keywords={break, case, catch, continue, debugger, default, delete, do, else,
  finally, for, function, if, in, instanceof, new, return, switch, this, throw,
  try, typeof, var, void, while, with, let},
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  morestring=[b]',
  morestring=[b]",
  sensitive=true
}

@@

\title{What does it take to create with domain-appropriate tools?}
\subtitle{A case study on the ``OROM'' system}
\author{Joel Jakubovic}
\affiliation{University of Kent, Canterbury}

@@

  However, this didn't happen. Or at least, it is happening glacially
  slowly. The brute fact is that whenever you want to create software, you
  go to a text editor and figure out how to translate your design into
  that. The text editors, being software, were written with the help of
  previous text editors, and so on. It's undeniable that text editors have
  improved, even if you think it peaked with Emacs. We just don't seem
  able to go beyond them where it matters, such as visual domains
  ill-fitted to monospaced ASCII.

  Amdahl's Law generalises the following idea: even when you spend hours
  of effort doubling the performance of a component used 1\% of the time,
  your reward is a system overall improved by a mere 0.5\%. Now, text
  coding is certainly ubiquitous, the 99\% case in programming, so you
  might wonder where I'm going with this. Well, a small improvement to
  text editing, if adopted by everyone, certainly does have a massive
  \emph{intermediate} effect --- but this only \emph{matters} to the
  extent that text was helping the programmers in the first place. If my
  goal is to draw or animate pictures, or create a digital synth from a
  frequency spectrogram, then giving me the ability to auto-indent my SVG
  markup is rather underwhelming as a productivity increase, as it doesn't
  target the core of the enterprise that makes it so hard.

  My experience of coding, most of the projects requiring shapes (such as
  GUIs), leads me to conclude that no matter how much I improve my skill
  at a particular language, knowledge of libraries or even general coding
  ability, my predicament stays the same. Our basic method of creating
  software is optimised for an ever-diminishing proportion of the software
  we actually want to make; ill-optimised for the graphics, layout,
  interactivity and and basic physics --- more on this later --- that we
  usually require.

@@

  Surely the trick is to \emph{use} coding to build something \emph{better
  than it}. And then use that, to build something even better. But there
  is an enormous breadth and depth of philosophies here, along with many
  failed historical attempts to do better --- or at least, ones that
  failed to catch on. But even worse than this is that in my very
  \emph{language} here I am making the same mistake as the text editor ---
  speaking in unqualified terms of ``better'' and ``worse'' as if there
  really is a one-size-fits-all solution to software creation!

  Of course what we \emph{really} want is the ability for people to create
  \emph{in the way that they think is best} in their particular context
  --- to equip them to feasibly create the tools that suit them for the
  thing they want to make. And second-order tools that suit them for
  making the first-order tools, etc. It would do no good to replace
  text-imperialism with anything-else-imperialism, which is one
  interpretation of calls for alternatives. This dream goes beyond
  familiar senses of what constitutes a ``craft'', as far as a strong
  melding of tool and material.

  Certainly in the real world there is only one ``material'' --- matter
  --- and humanity as a whole does create ``higher-order'' tools. A
  (future) software craft, however, could be expected to give this power
  to \emph{individuals} instead of the community alone. The
  one-size-fits-all approach is an understandable consequence of few
  designers catering to many users --- perhaps due to the necessity to
  specialise in topics, languages, libraries, and tools. Their complexity
  is part of the general difficulty of software creation, exacerbated by
  the one-size-fits-all approach, closing the circle.

  What we need is some system that not only lets us create software in a
  way that is ``close to the problem domain'' as decided by the
  user-developer, but also can augment or change itself to adapt to a
  different ``way of creating''. Existing systems seem to only have one of
  these properties without the other. Smalltalk and LISP try to minimise
  arbitrary commitments of language \emph{semantics} to be adaptable, but
  their being textual languages is a fairly tough commitment to break out
  of. And it is not so hard to make a specific, \emph{hard-baked} visual
  or alternative programming tool --- but it is hard to make it
  re-programmable \emph{without} having to go back to \emph{its} textual
  source code.

  Again, if someone wants to type out pictures in ASCII, let them ---
  whether they do it for a challenge, or even if they find that more
  natural for themselves. But equally, if I want to do it another way,
  then please give me that affordance. This is how I segue into the
  software artefact for which I have been attempting to build a natural
  representation. True, it is a programmer's artefact, but it is still
  representative of what any normal person has to do, insofar as:

@@

@@

  When I first read the paper Open, Reusable Object Models, I was hooked
  on its idea of a small but expressive starting system that could be
  self-improved into anything. It describes a late-bound\footnote{Fewer
    commitments; more things determined by runtime conditions},
  Smalltalk-style\footnote{OOP with more emphasis on object instances and
    messaging (methods) as in a distributed system; less emphasis on class
    hierarchies implementing traditional data structures} objects and
  messaging environment that the authors call ``Id'' --- but which I refer
  to as OROM, for pronunciation and Googleability.

  In summary, an object is a key-value mappings of state which can change
  as a result of messages received by it. A message is sent by first
  \emph{bind}-ing its name to its implementation, i.e.~specific code which
  is then run in the context of the receiver. ``Bind'' is accomplished via
  another message send, this time to the receiver's \emph{vtable}, or
  ``behaviour''. This is another object whose state defines a particular
  way of associating ``message name'' to ``implementation code''. This
  continues up the vtable-chain, and terminates at a base case. The system
  as a whole is ``bootstrapped'' into existence by its initialisation
  code. Each step of this code makes use of any parts of the system set up
  by the previous steps.

  The paper itself consists of mostly prose, several code listings, and
  full C sources for a sample implementation at the end. It also provides
  several diagrams. But to understand it, I repeatedly found myself
  drawing \emph{extra} diagrams. For example, the first acts of the
  running system boil down to initialising the three or so objects. This
  consists of allocating memory, interpreting it as a C struct and then
  filling in fields in a mundane manner. I had great difficulty following
  the specifics in my head, but when I drew tables in the style of their
  diagrams I readily saw what what going on. I personally would have
  preferred these to have been in the paper in the first place, but I am
  not necessarily representative of those who read it. The
  ``one-size-fits-all'' approach is perhaps unavoidable for static media
  like print.

  Other areas like messaging semantics could still be quite confusing.
  After all, should we expect to be able to predict the entire future
  evolution of a dynamical system from a static description of its initial
  state, such as source code? Is this not why debuggers exist? Having
  ``de-compiled'' English text and C source code into object diagrams, it
  is a shame to have to compile it all back to struct member assignments.

@@

  But further than that --- why can't the system be \emph{built} out of
  tables in the first place? Shouldn't this be the main takeaway from the
  amount of time we spend prototyping, explaining and designing software
  as diagrams on paper? Why must the ``natural representation'' be
  restricted to the finished product?

@@

  \hypertarget{orom-as-html-tables}{%
  \subsection{OROM as HTML tables}\label{orom-as-html-tables}}

@@

  I was grateful for the browser's management of graphical layout,
  resizable text fields, and keeping the DOM tree synchronised with what
  you actually see. This last property enabled me to make the decision to
  \emph{directly} encode much of the system state in the DOM, achieving
  basic liveness (the thing on the screen is the actual thing) for the
  keys and values of obj-dicts. For each obj-dict I used a two-column
  \texttt{\textless{}table\textgreater{}} within a
  \texttt{\textless{}div\textgreater{}}, matching the diagrams, and gave
  the rows CSS class names matching the keys for easy lookup. Sending
  messages relied on the JavaScript console, but existing values in text
  boxes (such as method implementations) could be edited directly. Also
  thanks to the browser, I could easily make text boxes and
  \texttt{\textless{}div\textgreater{}}s resizable as necessary.

  However, the choice of ordinary HTML as a substrate proved rather
  two-edged. The browser requires many features to do its job of rendering
  complex web pages. And sadly, as its client, I could only make use of
  those capabilities which the W3C had decided, at the time of authorship,
  were worth the effort exposing in JavaScript. For anything else, the
  browser is a black box, and this was very frustrating in the following
  case.

  \hypertarget{the-radical-concept-of-arrows-that-stay-on-the-shapes}{%
  \subsubsection{The Radical Concept of Arrows That Stay On The
  Shapes}\label{the-radical-concept-of-arrows-that-stay-on-the-shapes}}

@@

  Reluctantly, I stuck with my plan B: each object has a numerical ID and
  pointers are just fields containing a number, followed using a
  \texttt{deref()} function.

@@

  \hypertarget{orom-as-files-and-directories}{%
  \subsection{OROM as files and
  directories}\label{orom-as-files-and-directories}}

@@

  \hypertarget{orom-as-svg-trees}{%
  \subsection{OROM as SVG trees}\label{orom-as-svg-trees}}

@@

  \hypertarget{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}{%
  \section{Inevitable Requirements Of Most Software (and the work we
  must do to meet
  them)}\label{inevitable-requirements-of-most-software-and-the-work-we-must-do-to-meet-them}}

@@

  \hypertarget{as-a-mere-consumer}{%
  \subsection{As A ``Mere Consumer''}\label{as-a-mere-consumer}}

@@

  \hypertarget{retained-mode-vector-graphics}{%
  \subsubsection{Retained-Mode Vector
  Graphics}\label{retained-mode-vector-graphics}}

@@

  On this requirement, SVG fits the bill very well. Although it is not
  part of the JavaScript language \emph{per se}, it is a standard and
  widely supported technology of the Web \emph{platform}. We can observe
  that anyone with a browser \emph{in principle} has access to a powerful
  vector graphics editor --- just one with no GUI. I will return to this
  later.

@@

  \hypertarget{basic-assumptions-about-physical-objects}{%
  \subsubsection{Basic Assumptions About Physical
  Objects}\label{basic-assumptions-about-physical-objects}}

@@

  All humans learn a basic set of expectations about the things they see
  around them. Some of these, such as ``things fall down'', are not
  generally appropriate to software UIs --- perhaps because the screen has
  a role in our lives a more of a table work surface rather than a
  vertical wall, even if it is vertical in real life.

@@

  \hypertarget{maintaining-relationships-over-time}{%
  \subsubsection{Maintaining Relationships Over
  Time}\label{maintaining-relationships-over-time}}

@@

  \hypertarget{nut-cracking-with-sledgehammers}{%
  \subsubsection{Nut-Cracking With
  Sledgehammers}\label{nut-cracking-with-sledgehammers}}

@@

  The inevitable requirements I suggest here, do \emph{not} necessitate
  Fully General anything. In fact, such representations would make it more
  cumbersome to express what I wanted in OROM. As the old wisdom goes,
  there's no point expressing a simple regex search as an arbitrary Turing
  machine; my boxes don't \emph{have} a moment of inertia or a mass and I
  don't \emph{need} a linear optimisation solver for my space management
  --- for the time being.

@@

  \hypertarget{as-a-developer}{%
  \subsection{As A ``Developer''}\label{as-a-developer}}

@@

  \hypertarget{practicality-vs.-high-aesthetics}{%
  \subsubsection{Practicality vs.~High
  Aesthetics}\label{practicality-vs.-high-aesthetics}}

@@

  \begin{verbatim}
  Entity.prototype.addState = function(key) {
    let tr = document.createElement('tr');
    tr.setAttribute('class', key);

    let td = document.createElement('td');
    td.setAttribute('class', 'key');
    td.textContent = key;
    tr.appendChild(td);
    ...
  \end{verbatim}

  and in OROM/SVG the story is similar. It takes advantage of the powerful
  developer tools included in browsers (e.g.~the DOM inspector), and
  avoids the need to manually synchronise separate Model and View for such
  encoded state. With a little work, it also allowed the system a large
  degree of externalisability --- see Section ??. I consider this as
  practical, expedient, ``hijacking'' of these substrates because it
  either goes against their intended use (HTML tables for presentation
  rather than semantics) or are simply ``odd'' (using certain arrangements
  of shapes to ``encode'' machine-readable semantics).

@@

  Another example of practicality was in the issue of positioning and
  sizing. In Section ?? I mentioned that I was able to sidestep some
  layout work by offloading it to the user. While building the system I
  tended to want to draw boxes into existence, and for that my brain's
  existing aesthetic algorithms can decide where they go. There are areas
  where the machine must be able to figure this out itself --- I don't
  want to be constantly bothered by prompts to place boxes as they are
  allocated by some running code --- but for one-offs it is a very
  convenient way to do a little less yak-shearing.

@@

  \hypertarget{positioning-and-sizing}{%
  \subsubsection{Positioning and Sizing}\label{positioning-and-sizing}}

@@

  \hypertarget{visible-coordinate-systems}{%
  \subsubsection{Visible Coordinate
  Systems}\label{visible-coordinate-systems}}

@@

  All child elements of a node transform with it, so already SVG has baked
  in a basic facility for translational rigidity. This is only available
  as a tree hierarchy, but it is still useful. However, it conflicts with
  my early decision to have Point objects all share the global co-ordinate
  system. This way, connecting ``this point'' to the mouse pointer, etc.
  is not infuriating to express. Still, it was necessary in the case of
  certain elements --- especially those which must transcend the tree
  structure altogether, like arrows between boxes --- to bite this bullet,
  one way or another.

@@

  Again, I return to how we tend to work things out in the freedom of
  paper. Co-ordinate systems, here merely positionally displaced, have
  their origins here and there and have vectors between them. The rods
  thus far let me visually express relations between global Points; now
  was a question of expressing one global Point as a displacement from
  another (the \texttt{\textless{}g\textgreater{}} transform). New rod
  Observables \texttt{p2\_from\_p1} and \texttt{p1\_from\_p2} do the
  vector subtraction, which can then be propagated as local co-ordinates
  to children. It is nicer to express the relation (as well as see it!)
  this way --- see Figure ?? above.

  \hypertarget{context-appropriate-ontologies}{%
  \subsubsection{Context-appropriate
  ontologies}\label{context-appropriate-ontologies}}

  Each API has its own conventions, including a way of naming and
  structuring expressions --- an ontology. The one-size-fits-all approach
  is exemplified in such interfaces. For example, in SVG we express a
  rectangle by
  \texttt{\textless{}rect\ x="10"\ y="10"\ width="600"\ height="400"\textgreater{}}.
  The SVG specification \emph{defines} to the user that a rect \emph{is} a
  top-left corner, a width, and a height --- and that's it. However, this
  SVG-approved parametrisation of a rectangle is far from the only
  one(1){[}https://www.shift-society.org/salon/papers/2018/critiques/critique-semprola.pdf{]},
  and thus is, unsurprisingly, ill-fitted to some contexts.

  For example, I find it natural to resize boxes by dragging any of their
  four corners, so I wanted this in OROM/SVG. In this context, a
  ``rectangle'' is \emph{seen as} four points: top-left, top-right,
  bottom-left, bottom-right. Obviously this is not a \emph{minimal}
  description, since given e.g.~the top-left and bottom-right, the other
  two points can be inferred. But one way or another, to be able to drag
  any of them, all four points must be present at \emph{some} level. This
  alternative ontology was polyfilled in the form of a ``rect controls''
  class that can be attached manually to any SVG
  \texttt{\textless{}rect\textgreater{}}. The SVG \texttt{x} and
  \texttt{y} are subscribed to the top-left Point; \texttt{width} and
  \texttt{height} are subscribed to Rod lengths.

  Another example is to be found in the DOM's event listener model.
  Conceptually, many ``events'' are in fact changes to the state of some
  physical (or conceptual) device. Some of these genuinely are different.
  But both keyboard keys and mouse buttons, for example, have two states
  --- pressed, not pressed --- which ought to make them interchangeable to
  an extent. This is essentially the logic behind ``key remapping'' in
  many PC games. But the official ontology of the Web makes it nontrivial
  to do this.

  To start with, the situation is modelled not as a time-varying value
  changing, but instead as a sort of Cartesian product of event listeners.
  Rather than, say, a \texttt{left\_mouse\_button} piece of live-state, we
  get \texttt{onmousedown} and \texttt{onmouseup}. Thankfully we do not
  have \texttt{onAdown}/\texttt{onAup}, \texttt{onBdown/onBup}, \ldots{}
  all the way to \texttt{onZdown}/\texttt{onZup}, just
  \texttt{onkeydown}/\texttt{onkeyup}. But this is just one of the many
  possible ways to slice this 3-dimensional\footnote{Device (mouse,
    keyboard), sub-device (button, key), state (up, down)} space.

  In OROM/SVG I did not quite want to re-map keys, but I did often want to
  have things follow the mouse when dragged. In order to do this, I
  reified the mouse pointer and its position, letting me write
  \texttt{subscribe(point.position,\ pointer.position)}. The Point has an
  \texttt{is-considering-me?} Observable wrapping
  \texttt{onmouseover}/\texttt{onmouseout}, and the LMB\footnote{left
    mouse button} is reified into \texttt{left\_mouse\_button\_is\_down}
  to be explicit. The aforementioned subscription is set up whenever
  \texttt{is-considering-me?} and \texttt{left\_mouse\_button\_is\_down}
  become true, and torn down otherwise. I tended to think of this in the
  form ``subscribe to pointer \emph{only when} pointer is-considering-me
  \emph{and} LMB is down'', but I could live with this as a future
  polyfill in JavaScript.

  It takes some frustration and experience to get used to the idea that
  you have a right to polyfill in an alternative representation. Before I
  came to this conclusion, I used to twist my head around translating my
  intention into the \texttt{x,y,width,height} parameters, and
  un-translating when reading back the code I had produced. Once you get
  used to having to adapt your mental imagery to a single way of doing
  things (i.e.~learning to code), it makes sense to simply expect to see
  more of it --- especially when you know you are new, surrounded by
  veterans who see no problem, etc. But nowadays I take the position that
  these are simply \emph{widespread} failings of our way of doing things,
  instead of \emph{our} failure to adapt to the way software is.

  What would it look like to support multiple ontologies? There are the
  two answers: anticipate the possibilities ahead of time, or support
  users adding their own. The latter seems to imply polyfilling. As for
  the former: in general, anticipating the diversity of ways someone might
  look at the world is doomed to fail. But in the case of fairly
  \emph{formalised} concepts such as geometrical shapes or mathematics, I
  could suggest that simple \emph{under-specification} is the root of the
  problem in SVG. Instead of explaining in English that ``\texttt{x} and
  \texttt{y} are the co-ordinates of the top left-hand corner\ldots{}'',
  it might be better to make these relations machine-readable or
  \emph{embodied} in the API. For the rect this might look like:

  A \texttt{rect} is\ldots{} - a \texttt{polygon} (defined elsewhere) -
  defined by 4 \emph{degrees of freedom} \texttt{x\ y\ width\ height}
  (internal representation) - where there are 4 \emph{vertices}, all
  \texttt{point}s + \texttt{{[}x,y{]}} ``top-left'' +
  \texttt{{[}x+width,y{]}} ``top-right'' + \texttt{{[}x,y+height{]}}
  ``bot-left'' + \texttt{{[}x+width,y+height{]}} ``bot-right''

  The hope is that if we then specify enough information --- say, the
  bot-left and top-right --- then the runtime has enough information to
  derive its internal 4 degrees of freedom. In a crude sense I have
  successfully anticipated the most obvious ontologies of a rectangle
  here. However, I missed out the ``centre plus half-width and
  half-height'' formulation, among many others.

  Is this a futile effort even for precise mathematical knowledge
  structures, or could enough formalisation of Euclidean geometry in the
  Web platform put an end to this sort of polyfilling?\footnote{This
    relates to my belief that type systems and other auto-reasoning tools
    are trapped in a doomed quest for ``closed-form'' AI, representable as
    a LaTeX formula.}

  \hypertarget{extensional-functions}{%
  \subsubsection{Extensional Functions}\label{extensional-functions}}

  Time and time again we come across the same pattern of partitioning
  system state: trees or graphs of dictonaries, a.k.a. Maps, a.k.a.
  associative arrays. I am talking about filesystem paths
  \texttt{/path/to/some/file}, Python / Java modules
  \texttt{com.example.pkg.subpkg}, JavaScript objects
  \texttt{window.my\_obj.component}. The common case is an association of
  a textual (string) name, to an arbitrary value. I find it useful to see
  this as a mathematical ``function'' defined \emph{extensionally} by
  listing its input/output mappings -- this opposed to an
  \emph{intensional} definition such as
  \texttt{x\ \textbackslash{}mapsto\ 2x+3}, or a computer program.

  Extensional functions are perhaps the most basic form of Knowledge
  Representation, and match natural language very well. ``The bicycle's
  wheels' spokes are silver'' straightforwardly translates to a function
  equation
  \texttt{root\ (bicycle)\ (wheels)\ (spokes)\ (colour)\ =\ root\ (silver)}.
  That is, whatever object is the output of \texttt{silver} in the
  top-level \texttt{root} function, the output of \texttt{colour} (in the
  function on the left) points to the same object. The ability to
  partition a system in this way enables what
  (2){[}https://github.com/amb26/papers/blob/master/ppig-2016a/ppig2016a.pdf{]}
  calls a ``natural co-ordinate system'' for a piece of software, crucial
  for understanding and adaptability by others (so long as it is
  externalised).

  It seems that this way of expressing the ``parts'' of a system is an
  inevitable requirement of any programming substrate. Some languages,
  such as C, do have static, \emph{compile-time} associative arrays
  (\texttt{struct}s). In my experience this is not enough, and it's
  necessary to bring in a library or clutter the code with a home-grown
  approximation. Some parts of the OROM authors' C code were confusing
  until I realised they were just the guts of a basic associative-array
  implementation; when I switched to JavaScript, these lines vanished.
  Perhaps another strength of JavaScript is its low concrete syntax cost
  for \emph{instances} or \emph{literals} of associative arrays. Writing
  or reading

  \begin{lstlisting}[language=JavaScript]
  a = {
    b1: { c1: z, c2: y },
    b2: { c3: x, c4: w }
  };
  \end{lstlisting}

  is more WYSIWYG\footnote{What You See Is What You Get} than the
  imperative-style

  \begin{lstlisting}[language=JavaScript]
  a = new Map();
  a.set('b1', new Map());
  a.get('b1').set('c1', z);
  a.get('b1').set('c2', y);
  a.set('b2', new Map());
  a.get('b2').set('c3', x);
  a.get('b2').set('c4', w);
  \end{lstlisting}

  The latter is unfortunately still required in JavaScript for extensional
  functions with \emph{non-string} inputs.

  The decision to switch from the flat tabular representation in OROM/HTML
  to the nestable, Boxer-like structure in OROM/SVG permitted extensional
  (tree) functions in its substrate. True, I could always encode these in
  flat tables by having them reference each other. But this is like the
  second listing above: hard to follow. In the case of distinguishing
  between ordinary state mappings and the ``method dictionary'' of
  vtables, I actually stored method mappings with a ``-'' character to
  avoid doing this.

  (Screenshot showing this)

  In OROM/SVG, I can directly express the model I was thinking of: the
  vtable box has a \texttt{methods} box and that's where the method boxes
  go.

  (Screenshot showing this)

  \hypertarget{externalisability}{%
  \subsubsection{Externalisability}\label{externalisability}}

  This refers to exposing structured program state to the user. This data
  can then be saved and used to restore the system at a later date, but it
  can also be tweaked with corresponding changes reflected in the system.

  Externalisability, or its weaker cousin \emph{persistence}, was
  absolutely necessary to continue OROM/SVG development past a certain
  point.
